cardinalR: Generating interesting high-dimensional data structures

### 2025-09-05 paper-cardinalR.tex
Rendered after Michael's changes

There are numerous packages available in R for generating synthetic data, each designed with unique characteristics and focus areas. For example, \texttt{geozoo} (\citet{barret2016}) offers a large collection of geometric objects, allowing users to create and analyze specific shapes, primarily in lower-dimensional spaces. Another useful package is \texttt{snedata} (\citet{james2025}), which provides tools for generating simplified datasets useful for evaluating dimensionality reduction techniques like tSNE, often focusing on understanding and evaluating low-dimensional embeddings of complex data structures. Additionally, \texttt{mlbench} (\citet{friedrich2024}) includes a collection of well-known benchmark datasets commonly associated with established classification or regression challenges. In the field of single-cell omics, \texttt{splatter} (\citet{luke2017}) is particularly simulate complex biological data, effectively capturing nuances such as batch effects and differential expression.

There is a valuable opportunity to improve the generation of high-dimensional data structures by integrating geometric principles with advanced noise control and customizable clustering. The \texttt{geozoo} package provides a strong foundation but could be enhanced to support high-dimensional extensions with controlled noise and user-defined parameters for clustering. Similarly, while \texttt{snedata} focuses on abstract datasets for dimensionality reduction, adding features for generating high-dimensional data from geometric layouts would enhance its usability. The \texttt{mlbench} package could also benefit from allowing users to create datasets with specific geometric structures and noise profiles. Additionally, although \texttt{splatter} specializes in biological data simulation, it could be expanded to offer a broader framework for generating diverse geometric structures across dimensions, enabling detailed control over noise and clustering. Addressing these areas could lead to more robust high-dimensional data generation tools.

To address this gap, this paper introduces the \texttt{cardinalR} R package. This package provides a collection of functions designed to generate customizable data structures in any number of dimensions, starting from basic geometric shapes. \texttt{cardinalR} offers important functionalities that extend beyond the capabilities of existing tools, allowing users to: (i) construct high-dimensional datasets based on geometric shapes, including the option to enhance dimensionality by adding controlled noise dimensions; (ii) introduce adjustable levels of background noise to these structures; and (iii) create complex clustered data arrangements by using fundamental geometric forms, while maintaining their positions, scales, orientations, and sample sizes in arbitrary dimensional spaces. By providing these integrated features, \texttt{cardinalR} aims to provide researchers to generate more explainable and challenging synthetic datasets focused to the specific needs of evaluating algorithms in high-dimensions. This bridges the gap between geometric foundations and the flexible generation of complex synthetic data.

@@

A branching structure (Figure \ref{fig:branch-proj}) captures trajectories that diverge or bifurcate from a common origin, similar processes such as cell differentiation in biology. We introduce a set of data generation functions specifically designed to simulate high-dimensional branching structures with various geometry, number of points, and the number of branches. Table \ref{tab:branching-tb-pdf} outlines these functions. The main arguments of the functions described in Table \ref{tab:arg-branching-tb-pdf}.

@@

Each branch \(i\) is constructed using \(X_1 \sim U(-2, 2)\) and \(X_2 = \exp(\pm s_iX_1) + \epsilon\), where: \(\epsilon \sim U(0, 0.1)\) adds local jitter, \(s_i \in [0.5, 2]\) is randomly sampled for each branch. The sign alternates between branches (odd: negative, even: positive exponent). This creates mirror-symmetric branches with different steepness and curvature where odd-numbered branches: decay pattern (reflected) and even-numbered branches: growth pattern. For \(p > 2\), Gaussian noise \(X_j \sim N(0, 0.1^2)\) is added to embed the \(2\text{-}D\) branches into \(p\text{-}D\), where \(j = 3, \dots, p\).

@@

Each branch is a segment of a line with added jitter to simulate measurement noise. The branches differ in direction and location. Branches \(1\) and \(2\) are initialized with fixed slopes and intercepts. The Branch \(1\) is generated from \(X_1 \sim U(-2, 8)\), \(X_2 = 0.5X_1 + \epsilon\), where \(\epsilon \sim U(0, 0.5)\). The Branch \(2\) is generated from \(X_1 \sim U(-6, 2)\), \(X_2 = -0.5X_1 + \epsilon\), where \(\epsilon \sim U(0, 0.5)\). Branches \(3\) to \(k\) are added iteratively. Each additional branch \(i\) starts at a location outside predefined exclusion zones to avoid overlap with the initial two branches. The \(X_1\) values are defined over a short range, from \(x_{start}\) to \(x_{start} + 1\). The \(X_2\) value is calculated using the formula \(X_2 = s_i(X_1 - x_{start}) + y_{start} + \epsilon\), where \(s_i\) is a chosen slope from a selected branch, and \(\epsilon \sim U(0, 0.2)\). For \(p > 2\), Gaussian noise \(X_j \sim N(0, 0.05^2)\) is added to embed the \(2\text{-}D\) branches into \(p\text{-}D\), where \(j = 3, \dots, p\).

>>>>

There are numerous packages available in R for generating synthetic data, each designed with unique characteristics and focus areas.The \texttt{geozoo} package (\citet{barret2016}) offers a large collection of geometric objects, allowing users to create and analyze specific shapes, primarily in lower-dimensional spaces. Another useful package is \texttt{snedata} (\citet{james2025}), which provides tools for generating simplified datasets useful for evaluating dimensionality reduction techniques like tSNE, often focusing on understanding and evaluating low-dimensional embeddings of complex data structures. Additionally, \texttt{mlbench} (\citet{friedrich2024}) includes a collection of well-known benchmark datasets commonly associated with established classification or regression challenges. In the field of single-cell omics, \texttt{splatter} (\citet{luke2017}) is designed to simulate complex biological data, capturing field-specific nuances such as batch effects and differential expression.

There is an opportunity to improve the generation of high-dimensional data structures by integrating geometric principles with advanced noise control and customizable clustering. The \texttt{geozoo} package provides a strong foundation but could be enhanced to support high-dimensional extensions with controlled noise and user-defined parameters for clustering. Similarly, while \texttt{snedata} focuses on abstract datasets for dimensionality reduction, adding features for generating high-dimensional data from geometric layouts would enhance its usability. The \texttt{mlbench} package could also benefit from allowing users to create datasets with specific geometric structures and noise profiles. Additionally, although \texttt{splatter} specializes in biological data simulation, it could be expanded to offer a broader framework for generating diverse geometric structures across dimensions, enabling detailed control over noise and clustering. Addressing these areas could lead to more robust high-dimensional data generation tools.

To address this gap, this paper introduces the \texttt{cardinalR} R package. This package provides a collection of functions designed to generate customizable data structures in any number of dimensions, starting from basic geometric shapes. \texttt{cardinalR} offers important functionalities that extend beyond the capabilities of existing tools, allowing users to: (i) construct high-dimensional datasets based on geometric shapes, including the option to enhance dimensionality by adding controlled noise dimensions; (ii) introduce adjustable levels of background noise to these structures; and (iii) combine high-dimensional datasets into a single multi-faceted, clustered dataset in a space of arbitrary dimension. By providing these features, \texttt{cardinalR} provides researchers with a method to generate more explainable and challenging synthetic datasets focused to the specific needs of evaluating algorithms in high-dimensions. This bridges the gap between geometric foundations and the flexible generation of complex synthetic data.

@@

A branching structure (Figure \ref{fig:branch-proj}) captures trajectories that diverge or bifurcate from a common origin, similar to processes such as cell differentiation in biology. We introduce a set of data generation functions specifically designed to simulate high-dimensional branching structures with various geometry, number of points, and the number of branches. Table \ref{tab:branching-tb-pdf} outlines these functions. The main arguments of the functions described in Table \ref{tab:arg-branching-tb-pdf}.

@@

Each branch \(i\) is constructed using \(X_1 \sim U(-2, 2)\) and \(X_2 = \exp(\pm s_iX_1) + \epsilon\), where: \(\epsilon \sim U(0, 0.1)\) adds local jitter, \(s_i \in [0.5, 2]\) is randomly sampled for each branch. The sign alternates between branches (odd: negative, even: positive exponent). This creates mirror-symmetric branches with different steepness and curvature where odd-numbered branches represent a decay pattern (reflected) and even-numbered branches represent a growth pattern. For \(p > 2\), Gaussian noise \(X_j \sim N(0, 0.1^2)\) is added to embed the \(2\text{-}D\) branches into \(p\text{-}D\), where \(j = 3, \dots, p\).

@@

Each branch is a segment of a line with added jitter to simulate measurement noise. The branches differ in direction and location. Branches \(1\) and \(2\) are initialized with fixed slopes and intercepts. Branch \(1\) is generated from \(X_1 \sim U(-2, 8)\), \(X_2 = 0.5X_1 + \epsilon\), where \(\epsilon \sim U(0, 0.5)\). The Branch \(2\) is generated from \(X_1 \sim U(-6, 2)\), \(X_2 = -0.5X_1 + \epsilon\), where \(\epsilon \sim U(0, 0.5)\). Branches \(3\) to \(k\) are added iteratively. Each additional branch \(i\) starts at a location outside predefined exclusion zones to avoid overlap with the initial two branches. The \(X_1\) values are defined over a short range, from \(x_{start}\) to \(x_{start} + 1\). The \(X_2\) value is calculated using the formula \(X_2 = s_i(X_1 - x_{start}) + y_{start} + \epsilon\), where \(s_i\) is a chosen slope from a selected branch, and \(\epsilon \sim U(0, 0.2)\). For \(p > 2\), Gaussian noise \(X_j \sim N(0, 0.05^2)\) is added to embed the \(2\text{-}D\) branches into \(p\text{-}D\), where \(j = 3, \dots, p\).
